Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-03 Paper_Draft2/Draft2.tex
embedding all the images

This exploration aims to evaluate anti-Black algorithmic bias and the
differing effects of the COMPAS algorithm between white and Black
defendants; as such, we filter the data to only include individuals
whose race is listed as Caucasian or African-American. Our data
therefore has 5,723 rows with information on defendant race, age,
gender, prior crimes, and two-year recidivism rate. The average
African-American defendant in our dataset is 29 years old, male, and has
committed two prior crimes. The average Caucasian defendant in our
dataset is 35 years old, male, and has committed one prior crime. The
median defendant for both races do not have any juvenile convictions.

@@

\begin{longtable}[]{@{}lrrr@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived & Total \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived & Total \\
\midrule
\endhead
African American & 1661 & 1512 & 3173 \\
Caucasian & 822 & 1278 & 2100 \\
Total & 2483 & 2790 & 5273 \\
\bottomrule
\end{longtable}

@@

With our baseline model, we ran the four different group fairness
metrics we chose and compared the results (pictured in Table
\ref{tab:baseline metrics table}).

@@

Our confusion matrix for the baseline model indicates that the false
positive and false negative percentages are 16.68\% and 22.06\%, which
are the lowest values for this matrix. The highest value is the value of
true positives, at 35.41\%, showing that of the total number of
predictions, 35.41\% were correct predictions of recidivism.

\begin{longtable}[]{@{}llll@{}}
\caption{Results of Baseline Fairness Metric Analysis
\label{tab:baseline metrics table}}\tabularnewline
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
\midrule
\endfirsthead
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
\midrule
\endhead
Statistical Parity Difference & 0 & -0.14 & White Defendants \\
Disparate Impact Ratio & 1 & 0.47 & White Defendants \\
Average Odds Difference & 0 & -0.44 & White Defendants \\
Equal Opportunity Difference & 0 & -0.41 & White Defendants \\
\bottomrule
\end{longtable}

@@

After running the reweighing algorithm, our fairness metrics are -0.015
for statistical parity difference, 0.015 for equal opportunity
difference, 0.014 for average odds difference, and 0.98 for disparate
impact ratio. All of these values now fall within the margins of
fairness (see reweighing metrics plot). These values make sense because
the reweighing algorithm chooses different weights depending on whether
an attribute is protected. This does an effective job of eliminating the
bias, shown by each of the values in the graph being within the ``fair''
range of the particular fairness metric. Overall, these fairness metrics
show that the reweighing algorithm improves the bias in the COMPAS
algorithm.

The highest percentage in this confusion matrix is for the percentage of
true negatives, defendants who were predicted to not recidivate and
actually did not, at 36.24\%, 0.83\% higher than the true negatives in
the baseline model. This means that the reweighing model marginally
improved the baseline model accuracy in predicting people who did not
recidivate. However, the percentage of true positives, or those who were
predicted to recidivate and did recidivate, is 24.72\%, 1.13\% lower
than that of the baseline model. This means that the reweighing model
slightly lowered the accuracy of the baseline model in predicting people
who did recidivate. As a result, the overall reweighing model accuracy
remains about the same as the baseline model accuracy. The false
negative and false positive percentages, or the predictions that proved
to be incorrect, are 23.20\% and 15.85\%, respectively. The re-weighing
model increased the number of false negatives by 1.14\% and decreased
the number of false positives by 0.83\%. We are most concerned about the
false positives, which indicate defendants who are predicted to
recidivate and do not actually recidivate. While there is a slight
improvement in the percent of false positives, it is still relatively
high at 15.85\%, indicating that we should further precede with
additional bias mitigation techniques.

@@

Our fairness metrics after running the prejudice remover are: 0.367 for
statistical parity difference, 0.321 for equal opportunity difference,
0.344 for average odds difference, and 3.3 for disparate impact ratio.
Like the model performance metrics suggested, the prejudice remover
approach resulted in an increased benefit for Black defendants. As the
orange bars on these plots show, the values of the fairness metrics have
reversed from their baseline values. Now all four metrics suggest an
unfair advantage for Black defendants. Thus, this approach removed the
model's prejudice against Black people, but it did not result in a
``fair'' model.

The highest percentage in this confusion matrix is for false positives,
defendants who the model predicted to recidivate and actually did not,
at 37.76\%. This false positive percentage concerns us because we do not
want defendants who do not recidivate to have unfairly long sentences
due to their (incorrectly) predicted recidivism. The number of true
positives, defendants who the model predicted to recidivate and actually
recidivated, is lower at 23.88\%. The percentage of false negatives is
lower, whereas the true negative is low at 14.33\%. This indicates that
the model does not do a good job in predicting the defendants who don't
recidivate. It is very likely that it will incorrectly predict that
someone will recidivate.

@@

After running the calibrated equalized odds algorithm, our fairness
metrics are: 0.134 for statistical parity difference, 0.088 for equal
opportunity difference, 0.146 for average odds difference, and 1.155 for
disparate impact ratio. As we can see from the orange bars, the
calibrated equalized odds approach results in all four metrics now
suggesting a benefit for the originally unprivileged group, Black
defendants. Though the values for statistical parity difference and
average odds difference are slightly above the range of what is
considered fair, the margin is much smaller than the original margin
between the value and the range of fairness (as indicated by the blue
bars). Thus, the calibrated equalized odds approach successfully
counteracts the bias against Black defendants and results in a mostly
fair model.

The highest percentage in this confusion matrix is for true negatives,
defendants who the model predicted to not recidivate and actually did
not, at 50.04\%. The number of true positives, defendants who the model
predicted to recidivate and actually recidivated, is much lower at
3.11\%. The percentage of false negatives is somewhat high for this
baseline model, whereas the amount of false positives is very low at
2.05\%. This indicates that the model does a good job in predicting the
defendants who don't recidivate. It is very unlikely that it will
incorrectly predict that someone will recidivate. Though the accuracy
score for the model is 0.53, lower than the baseline, this model is much
better at identifying true negatives.

>>>>

This exploration aims to evaluate anti-Black algorithmic bias and the
differing effects of the COMPAS algorithm between white and Black
defendants; as such, we filter the data to only include individuals
whose race is listed as Caucasian or African-American. Our data
therefore has 5,723 rows with information on defendant race, age,
gender, prior crimes, and two-year recidivism rate (Figure
\ref{fig:table snip}). The distributions of age (Figure
\ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and
recidivism (Figure \ref{fig:recidivism mosaic} and Table
\ref{tab:recid table}) all vary by race. The average African-American
defendant in our dataset is 29 years old, male, and has committed two
prior crimes. The average Caucasian defendant in our dataset is 35 years
old, male, and has committed one prior crime. The average defendant for
both races does not have any juvenile convictions.

@@

\begin{longtable}[]{@{}lrrr@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived &
Total\tabularnewline
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived &
Total\tabularnewline
\midrule
\endhead
African American & 1661 & 1512 & 3173\tabularnewline
Caucasian & 822 & 1278 & 2100\tabularnewline
Total & 2483 & 2790 & 5273\tabularnewline
\bottomrule
\end{longtable}

@@

With our baseline model, we ran the four different group fairness
metrics we chose and compared the results (pictured in Figre
\ref{fig:baseline metrics}).

@@

Our confusion matrix for the baseline model (see Figure
\ref{fig:baseline matrix}) indicates that the false positive and false
negative percentages are 16.68\% and 22.06\%, which are the lowest
values for this matrix. The highest value is the value of true
positives, at 35.41\%, showing that of the total number of predictions,
35.41\% were correct predictions of recidivism.

{\centering \includegraphics[width=1\linewidth]{../images/baseline_metrics}

\caption{Fairness Metrics of Baseline Model. For each of the four fairness metrics, the green section represents the range of fair values. The blue bars indicate the values of each of the metrics for the baseline model.}\label{fig:baseline metrics}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{../images/baseline_matrix}

\caption{Confusion Matrix of the Baseline Model}\label{fig:baseline matrix}
\end{figure}

@@

After running the reweighing algorithm, our fairness metrics are -0.015
for statistical parity difference, 0.015 for equal opportunity
difference, 0.014 for average odds difference, and 0.98 for disparate
impact ratio. All of these values now fall within the margins of
fairness (see Figure \ref{fig:reweighing metrics}). These values make
sense because the reweighing algorithm chooses different weights
depending on whether an attribute is protected. This does an effective
job of eliminating the bias, shown by each of the values in the graph
being within the ``fair'' range of the particular fairness metric.
Overall, these fairness metrics show that the reweighing algorithm
improves the bias in the COMPAS algorithm.

{\centering \includegraphics[width=1\linewidth]{../images/reweighting_metrics}

\caption{Fairness Metrics of Reweighing Model. For each of the four fairness metrics, the orange bars indicate the values for the reweighing model.}\label{fig:reweighing metrics}
\end{figure}

The highest percentage in the reweighing confusion matrix (Figure
\ref{fig:reweighing matrix}) is for the percentage of true negatives,
defendants who were predicted to not recidivate and actually did not, at
36.24\%, 0.83\% higher than the true negatives in the baseline model.
This means that the reweighing model marginally improved the baseline
model accuracy in predicting people who did not recidivate. However, the
percentage of true positives, or those who were predicted to recidivate
and did recidivate, is 24.72\%, 1.13\% lower than that of the baseline
model. This means that the reweighing model slightly lowered the
accuracy of the baseline model in predicting people who did recidivate.
As a result, the overall reweighing model accuracy remains about the
same as the baseline model accuracy. The false negative and false
positive percentages, or the predictions that proved to be incorrect,
are 23.20\% and 15.85\%, respectively. The re-weighing model increased
the number of false negatives by 1.14\% and decreased the number of
false positives by 0.83\%. We are most concerned about the false
positives, which indicate defendants who are predicted to recidivate and
do not actually recidivate. While there is a slight improvement in the
percent of false positives, it is still relatively high at 15.85\%,
indicating that we should further precede with additional bias
mitigation techniques.

{\centering \includegraphics[width=1\linewidth]{../images/reweighing_matrix}

\caption{Confusion Matrix of the Reweighing Model}\label{fig:reweighing matrix}
\end{figure}

@@

Our fairness metrics after running the prejudice remover are 0.367 for
statistical parity difference, 0.321 for equal opportunity difference,
0.344 for average odds difference, and 3.3 for disparate impact ratio.
Like the model performance metrics suggested, the prejudice remover
approach resulted in an increased benefit for Black defendants. As the
orange bars on Figure \ref{fig:prejudice metrics} show, the values of
the fairness metrics have reversed from their baseline values. Now all
four metrics suggest an unfair advantage for Black defendants. Thus,
this approach removed the model's prejudice against Black people, but it
did not result in a ``fair'' model.

{\centering \includegraphics[width=1\linewidth]{../images/prejudice_remover_metrics}

\caption{Fairness Metrics of Prejudice Remover Model. For each of the four fairness metrics, the orange bars indicate the values for the prejudice remover model.}\label{fig:prejudice metrics}
\end{figure}

The highest percentage in this confusion matrix (Figure
\ref{fig:prejudice matrix}) is for false positives, defendants who the
model predicted to recidivate and actually did not, at 37.76\%. This
false positive percentage concerns us because we do not want defendants
who do not recidivate to have unfairly long sentences due to their
(incorrectly) predicted recidivism. The number of true positives,
defendants who the model predicted to recidivate and actually
recidivated, is lower at 23.88\%. The percentage of false negatives is
lower, whereas the true negative is low at 14.33\%. This indicates that
the model does not do a good job in predicting the defendants who don't
recidivate. It is very likely that it will incorrectly predict that
someone will recidivate.

{\centering \includegraphics[width=1\linewidth]{../images/prejudice_matrix}

\caption{Confusion Matrix of the Prejudice Remover Model}\label{fig:prejudice matrix}
\end{figure}

@@

After running the calibrated equalized odds algorithm, our fairness
metrics are 0.134 for statistical parity difference, 0.088 for equal
opportunity difference, 0.146 for average odds difference, and 1.155 for
disparate impact ratio. As we can see in Figure \ref{fig:ceo metrics},
the calibrated equalized odds approach results in all four metrics now
suggesting a benefit for the originally unprivileged group, Black
defendants. Though the values for statistical parity difference and
average odds difference are slightly above the range of what is
considered fair, the margin is much smaller than the original margin
between the value and the range of fairness (as indicated by the blue
bars). Thus, the calibrated equalized odds approach successfully
counteracts the bias against Black defendants and results in a mostly
fair model.

{\centering \includegraphics[width=1\linewidth]{../images/calibrated_equalized_odds_metrics}

\caption{Fairness Metrics of Calibrated Equalized Odds Model. For each of the four fairness metrics, the orange bars indicate the values for the calibrated equalized odds model.}\label{fig:ceo metrics}
\end{figure}

The highest percentage in this confusion matrix (Figure
\ref{fig:ceo matrix}) is for true negatives, defendants who the model
predicted to not recidivate and actually did not, at 50.04\%. The number
of true positives, defendants who the model predicted to recidivate and
actually recidivated, is much lower at 3.11\%. The percentage of false
negatives is somewhat high for this baseline model, whereas the amount
of false positives is very low at 2.05\%. This indicates that the model
does a good job in predicting the defendants who don't recidivate. It is
very unlikely that it will incorrectly predict that someone will
recidivate. Though the accuracy score for the model is 0.53, lower than
the baseline, this model is much better at identifying true negatives.

{\centering \includegraphics[width=1\linewidth]{../images/ceo_matrix}

\caption{Confusion Matrix of the Calibrated Equalized Odds Model}\label{fig:ceo matrix}
\end{figure}
